Astr 513: Final Project

### 2016-05-10 Final_Project/ms.tex
More or less finished results and conclusions

\section{Planck 2015 Results}
\label{sec:planck}
In 2009 the European Space Agency launched the \Planck satellite in order
to measure the Cosmic Microwave Background with unparalleled precision.
By measuring the small anisotropies in the temperature of the CMB across
the entire sky one calculates the angular power spectrum, which can be
modeled using first principles (see, e.g., \citet{dodelson2003}).
By using a likelihood method similar to our own, the authors of
\citet{planck2013} estimated 14 nuisance parameters and 
six cosmological parameters, including the geometrical parameters,
$\Omega_M$, $\Omega_\Lambda$ and $\Omega_k$, that we use in our
{\sna} analysis. The only prior used by \citet{planck2013}
on cosmological parameters was from the WMAP result \citep{wmap2011}
for the optical depth $\tau$. We do not use this parameter
in our {\sna} model. \citet{planck2013} report marginalized
posterior distributions for all parameters,
which we report in \autoref{tab:priors}.
\EW{I'm assuming this should be the \Planck priors table?}

@@

\subsection{The Pan-STARRS1 sample}
\label{sec:rest}
\EW{Should we mention the issue with the covariance matrix in this section or the next? Or not at all?}
\citet{rest2014} uses the first 1.5 year of the first telescope of 
the Panoramic Survey Telescope and Rapid Response System (Pan-STARRS1) 
and a compilation of low-redshift {\sna} studies. The low-redshift 
sample consists of {\snea} collected by Cal\'{a}n/Tololo (Hamuy et al. 1996a), 
CfA1 (Riess et al. 1999), CfA2 (Jha et al. 2006), CfA3 (Hicken et al. 2009a), 
CSP (Contreras et al. 2010), and CfA4 (Hicken et al. 2012). After 
three quality cuts, 113 spectroscopically confirmed {\snea} from 
Pan-STARRS1 ($0.03 < z < 0.65$) and 222 light curves from 197 
low-redshift {\snea} are used for cosmological analysis.

@@

The correlation matrix is calculated from the covariance matrix according to
%
\begin{equation}
  \label{eq:correlation}
  Corr_{i,j} = \frac{\mathbf{C}_{i,j}}{\sqrt{\mathbf{C}_{i,i}}\sqrt{\mathbf{C}_{j,j}}}.
\end{equation}
%
Note that the correlation of a variable with itself is 1 by definition.
The correlation matrices for both of our data sets can be seen below in
\autoref{fig:cmatrix_both}.
%
\begin{figure}
  \includegraphics[width=0.5\linewidth]{figures/Betoule_correlation.eps}
  \includegraphics[width=0.5\linewidth]{figures/Rest_correlation.eps}
  \caption{{\it Left}: Correlation matrix of $\mu_0(z)$ from the
    \citet{betoule2014} sample. {\it Right}: Correlation matrix of $\mu_0(z)$ from 
	the \citet{rest2014} sample. Very few correlations exist in 
        \citet{betoule2014} likely due to the averaging over redshift bins. 
	Significant correlation exists in \citet{rest2014} due to dependencies 
	on systematic variables as seen in \autoref{eq:RestCovMatrix}.}
  \label{fig:cmatrix_both}
\end{figure}

@@

Two cosmologies are of interest to us: the  $\Lambda$CDM 
universe and the $w$CDM universe. In the  $\Lambda$CDM 
universe, $w=-1$ and constraints are placed on $\Omega_M$ and $h$. 
The $w$CDM universe is a one-parameter extension of  $\Lambda$CDM 
in which $w$ is an arbitrary parameter for a constant dark energy 
equation of state. In $w$CDM, constraints are placed on $\Omega_M$, $w$, 
and $h$. In both cosmologies, we assume the universe is spatially flat 
($\Omega_k = 0$) and thus $\Omega_{\Lambda}=1-\Omega_M$. 
\EW{Added \textbf{we assume} to the last sentence because $\Lambda$CDM and $w$CDM do not necessarily assume a spatially flat universe}

\subsection{Likelihood}
\label{sec:likelihood}
Using the above model for the distance modulus allows us to write a
probability that a model can generate our data:
%
\begin{align}
  \label{eq:likelihood}
  P(\mu(z)|\mathrm{cosmo}) &= \frac{1}{\sqrt{2\pi|\mathbf{C}|}}\nonumber \\
  &\qquad{}\exp\left[-\frac{1}{2} \sum_{i,j=0}^{N_{samples}}\left(\mu(z) - \mu_{\mathrm{model}}(z)\right)_i 
    \mathbf{C}^{-1}_{i,j}\left(\mu(z) - \mu_{\mathrm{model}}(z)\right)_j\right]
\end{align}
%
where $\mathbf{C}$ is the full covariance matrix between all of the $N_{samples}$ 
in the data, and $\mathrm{cosmo}$ represents the cosmology that we are investigating
(e.g. flat $\Lambda$CDM, $w$CDM, etc.). 
This is not the full picture though, since we must employ
Bayes' Theorum in order to include any priors on our choice of 
parameters and recover the posterior distribution function of our
parameters given our data
%
\begin{equation}
  \label{eq:posterior}
  P(\mathrm{cosmo}|\mu(z)) = P_{prior}(\mathrm{cosmo}) P(\mu(z)|\mathrm{cosmo}).
\end{equation}
%
\EW{Is this equation missing the normalization constant?}
\TM{No it is general. The equation before is normalized properly, and this one 
  doesn't specify the prior, so it is normalized in that the prior hasn't been
  specified yet.}
For this work we investigate a combination of priors. \CosmoSIS is implemented
in a way that requires the user to provide bounds on the parameter space, which we
detail for each parameter in \autoref{tab:priors}.
Within those bounds, we investigated differences between using flat prior $P_{prior}=1$ on
all parameters simultaneous in addition to \Planck priors on all parameters.

\begin{table}
  \centering
  \caption{Bounds on cosmological parameters 
    entering \autoref{eq:posterior}. Our sampler requires
    bounds in order to both populate the parameter space with
    parallelized samplers and to calculate an appropriate width
    for the step size.}
  \label{tab:priors}
  \begin{tabular}{lll}
    Parameter & Description & Bounds \\ \hline
    $h$ & Hubble constant$/100$ & $[0,2]$ \\
    $\Omega_M$ & Matter fraction at $z=0$ & $[0,1]$ \\
    $w$ & Dark energy E.O.S. & $[-2,0]$\\
  \end{tabular}
\end{table}

@@

\section{Results}
\label{sec:results}
The most likely values and uncertainties of $\Omega_M$ and $h$ 
in a $\Lambda$CDM model are listed in \autoref{tab:lcdm}. Included are results from
using the \citet{betoule2014} and \citet{rest2014} data with and without the addition 
of \Planck priors. Additionally, the 68\% and 95\% confidence contours can be seen
in \autoref{fig:betoule_lcdm_corner} and \autoref{fig:rest_lcdm_corner}.

Similary, \autoref{tab:wcdm}, \autoref{fig:betoule_wcdm_corner}, and \autoref{fig:rest_wcdm_corner}
detail the most likely values and confidence intervals of $\Omega_M$, $h$, and $w$ 
in a $w$CDM model. An extra entry is given in \autoref{tab:wcdm} for just our prior, the 
\Planck likelihood. 
\LC{Not done yet}

%\LC{Not sure where to put why we couldn't do the full analysis of Rest's data. 
%  I'm putting it here for now.}
%A complete analysis using the data from \citet{rest2014} is not done 
%in our paper since there are several issues with the publicly released 
%data. It is not apparent which uncertainties belonged to the statistical 
%covariance matrix since only the systematic covariance matrix is clearly 
%presented. Moreover, intrinsic dispersion of the data and photometric errors 
%of the distances depend on nuisance parameters in addition to the covariances 
%between the fit parameters. Fine-tuning nuisance parameters places a significant 
%time restraint on our work and is beyond the scope of this paper. For these reasons, 
%we only present the results for \citet{rest2014} that do not include the priors 
%from \citet{planck2013}. 

@@

    ~                          & $\Omega_m$ & $h$ \\ \hline

@@

    ~                          & $\Omega_m$ & $h$ & $w$ \\ \hline

@@

  \label{fig:betoule_wcdm_prior_bestfit}

@@

\section{Conclusions}
\label{sec:conclusions}
\TM{Talk about how the inclusions if priors changes the analysis.
  State whether the priors are strong or uninformative.}

>>>>

\section{Planck 2015 Results}
\label{sec:planck}
In 2009 the European Space Agency launched the \Planck satellite in order
to measure the Cosmic Microwave Background with unparalleled precision.
By measuring the small anisotropies in the temperature of the CMB across
the entire sky one calculates the angular power spectrum, which can be
modeled using first principles (see, e.g., \citet{dodelson2003}).
By using a likelihood method similar to our own, the authors of
\citet{planck2013} estimated 14 nuisance parameters and 
six cosmological parameters, including the geometrical parameters,
$\Omega_M$, $\Omega_\Lambda$ and $\Omega_k$, that we use in our
{\sna} analysis. The only prior used by \citet{planck2013}
on cosmological parameters was from the WMAP result \citep{wmap2011}
for the optical depth $\tau$. We do not use this parameter
in our {\sna} model. \citet{planck2013} report marginalized
posterior distributions for all parameters,
which we report in \autoref{tab:wcdm}.

@@

\subsection{The Pan-STARRS1 sample}
\label{sec:rest}
\citet{rest2014} uses the first 1.5 year of the first telescope of 
the Panoramic Survey Telescope and Rapid Response System (Pan-STARRS1) 
and a compilation of low-redshift {\sna} studies. The low-redshift 
sample consists of {\snea} collected by Cal\'{a}n/Tololo (Hamuy et al. 1996a), 
CfA1 (Riess et al. 1999), CfA2 (Jha et al. 2006), CfA3 (Hicken et al. 2009a), 
CSP (Contreras et al. 2010), and CfA4 (Hicken et al. 2012). After 
three quality cuts, 113 spectroscopically confirmed {\snea} from 
Pan-STARRS1 ($0.03 < z < 0.65$) and 222 light curves from 197 
low-redshift {\snea} are used for cosmological analysis.

@@

The correlation matrix is calculated from the covariance matrix according to
%
\begin{equation}
  \label{eq:correlation}
  Corr_{i,j} = \frac{\mathbf{C}_{i,j}}{\sqrt{\mathbf{C}_{i,i}}\sqrt{\mathbf{C}_{j,j}}}.
\end{equation}
%
Note that the correlation of a variable with itself is 1 by definition.
The correlation matrices for both of our data sets can be seen in
\autoref{fig:cmatrix_both}.
%
\begin{figure}
  \includegraphics[width=0.5\linewidth]{figures/Betoule_correlation.eps}
  \includegraphics[width=0.5\linewidth]{figures/Rest_correlation.eps}
  \caption{{\it Left}: Correlation matrix of $\mu_0(z)$ from the
    \citet{betoule2014} sample. {\it Right}: Correlation matrix of $\mu_0(z)$ from 
	the \citet{rest2014} sample. Very few correlations exist in 
        \citet{betoule2014} likely due to the averaging over redshift bins. 
	Significant correlation exists in \citet{rest2014} due to dependencies 
	on systematic variables as seen in \autoref{eq:RestCovMatrix}.}
  \label{fig:cmatrix_both}
\end{figure}

@@

Two cosmologies are of interest to us: the  $\Lambda$CDM 
universe and the $w$CDM universe. In the  $\Lambda$CDM 
universe, $w=-1$ and constraints are placed on $\Omega_M$ and $h$. 
The $w$CDM universe is a one-parameter extension of  $\Lambda$CDM 
in which $w$ is an arbitrary parameter for a constant dark energy 
equation of state. In $w$CDM, constraints are placed on $\Omega_M$, $w$, 
and $h$. In both cosmologies, we assume the universe is spatially flat 
($\Omega_k = 0$) and thus $\Omega_{\Lambda}=1-\Omega_M$.

\subsection{Likelihood}
\label{sec:likelihood}
Using the above model for the distance modulus allows us to write a
probability that a model can generate our data:
%
\begin{align}
  \label{eq:likelihood}
  P(\mu(z)|\mathrm{cosmo}) &= \frac{1}{\sqrt{2\pi|\mathbf{C}|}}\nonumber \\
  &\qquad{}\exp\left[-\frac{1}{2} \sum_{i,j=0}^{N_{samples}}\left(\mu(z) - \mu_{\mathrm{model}}(z)\right)_i 
    \mathbf{C}^{-1}_{i,j}\left(\mu(z) - \mu_{\mathrm{model}}(z)\right)_j\right]
\end{align}
%
where $\mathbf{C}$ is the full covariance matrix between all of the $N_{samples}$ 
in the data, and $\mathrm{cosmo}$ represents the cosmology that we are investigating
(e.g. flat $\Lambda$CDM, $w$CDM, etc.). 
This is not the full picture though, since we must employ
Bayes' Theorem in order to include any priors on our choice of 
parameters and recover the posterior distribution function of our
parameters given our data
%
\begin{equation}
  \label{eq:posterior}
  P(\mathrm{cosmo}|\mu(z)) = P_{prior}(\mathrm{cosmo}) P(\mu(z)|\mathrm{cosmo}).
\end{equation}
%
For this work we investigate a combination of priors. \CosmoSIS is implemented
in a way that requires the user to provide bounds on the parameter space, which we
detail for each parameter in \autoref{tab:priors}.
Within those bounds, we investigated differences between using flat prior $P_{prior}=1$ on
all parameters simultaneous in addition to \Planck priors on all parameters.

\begin{table}
  \centering
  \begin{tabular}{lll}
    Parameter & Description & Bounds \\ \hline
    $h$ & Hubble constant$/100$ & $[0,2]$ \\
    $\Omega_M$ & Matter fraction at $z=0$ & $[0,1]$ \\
    $w$ & Dark energy E.O.S. & $[-2,0]$\\
  \end{tabular}
  \caption{Bounds on cosmological parameters 
    entering \autoref{eq:posterior}. Our sampler requires
    bounds in order to both populate the parameter space with
    parallelized samplers and to calculate an appropriate width
    for the step size.}
  \label{tab:priors}
\end{table}

@@

\section{Results}
\label{sec:results}
The most likely values and uncertainties of $\Omega_M$ and $h$ 
in a $\Lambda$CDM model are listed in \autoref{tab:lcdm}. Included are results from
using the \citet{betoule2014} and \citet{rest2014} data with and without the addition 
of \Planck priors. Additionally, the 68\% and 95\% confidence contours can be seen
in \autoref{fig:betoule_lcdm_corner} and \autoref{fig:rest_lcdm_corner}. Similary, \autoref{tab:wcdm}, 
\autoref{fig:betoule_wcdm_corner}, and \autoref{fig:rest_wcdm_corner}
detail the most likely values and confidence intervals of $\Omega_M$, $h$, and $w$ 
in a $w$CDM model. An extra entry is given in \autoref{tab:wcdm} for just the 
\Planck likelihood. The best distance modulus fits to the \citet{betoule2014} and \citet{rest2014} data
from our MCMC chains for a $w$CDM model with \Planck priors are shown in \autoref{fig:wcdm_prior_bestfit}.

A complete analysis using the data from \citet{rest2014} reveals  
several issues with their publicly released data. It is not apparent 
which uncertainties belonged to the statistical covariance matrix since 
only the systematic covariance matrix is clearly labeled in the released data. Moreover, intrinsic 
dispersion of the data and photometric errors of the distances depend on nuisance 
parameters in addition to the covariances between the fit parameters. Fine-tuning 
nuisance parameters is omitted by us since it places a significant time restraint on our work. Due to 
these issues, most of the values and contours calculated with the \citet{rest2014} data without 
\Planck priors do not give reasonable results. 

@@

    ~                          & $\Omega_M$ & $h$ \\ \hline

@@

    ~                          & $\Omega_M$ & $h$ & $w$ \\ \hline

@@

  \label{fig:wcdm_prior_bestfit}

@@

\section{Conclusions}
\label{sec:conclusions}
\LC{Not sure if I should say anything more about \autoref {fig:wcdm_prior_bestfit} here.}
As seen in \autoref{tab:lcdm} and \autoref{fig:betoule_lcdm_corner}, the addition of the \Planck priors
to the \citet{betoule2014} data tightens the constraints on $\Omega_M$ and $h$ in a $\Lambda$CDM model. 
Moreover, the \Planck priors vastly improve the constraint on $\Omega_M$ despite the issues in \citet{rest2014} data
as demonstrated in \autoref{fig:rest_lcdm_corner}.

Comparably, the inclusion of the \Planck priors to the \citet{betoule2014} data also tighten the 
constraints on $\Omega_M$, $h$, and $w$ in a $w$CDM model as seen in \autoref{tab:wcdm} and  
\autoref{fig:betoule_wcdm_corner}. This also applies to the \citet{rest2014} data as illustrated 
in \autoref{fig:rest_wcdm_corner}, where $\Omega_M$ and $w$ are well constrained compared to the case with a flat prior. 
The confidence contours and most likely values are nearly identical to the ones from the \Planck likelihood.

We remind the readers that \citet{planck2013} accounts for nuisance and cosmological
parameters unlike our analysis of the \citet{betoule2014} and \citet{rest2014} data, which disregards 
nuisance parameters. Nonetheless, we conclude that the inclusion of information from other cosmological 
probes as priors strengthens {\sna} inferences of cosmological parameters. Furthermore, \Planck priors prove
to be informative especially in the case of problematic data.
